Supporting Information for\\ ``A weakly structured stem for human origins in Africa''

### 2022-07-18 supplement.tex
nama threshold section

The four inferred demographic models (single origin, continuous migration,
merger without stem migration, and merger with stem migration) were converted
to \demes format (Gower, Ragsdale, \emph{et al.}, in prep) and used as
the input demography for \msprime simulations. We sampled the same
number of individuals from each population as used in the inference (Nama: 44,
Mende: 85, Gumuz: 23, combined Oromo/Amhara: 46, British: 91, and Vindija
Neanderthal: 1). We assumed a generation time of 29 years, a per-base mutation
rate of $1.25\times10^{-8}$, a per-base recombination rate of $1\times10^{-8}$
(corresponding to 1cM/Mb), and we simulated 100 replicates with sequence
lengths of 50Mb per replicate. We computed the conditional site frequency
spectrum (cSFS) conditioned on observing the derived variant in the Neanderthal
sample, and reconstructed gene genealogies using \Relate to obtain coalescence
rate trajectories and distributions of deep branches passing through the
Neanderthal branch. Each analysis followed the same procedure as described
above in the analysis of the real data.

@@

@@

>>>>

The four inferred demographic models (single origin, continuous migration,
merger without stem migration, and merger with stem migration) were converted
to \demes format \citep{Gower2022-yn} and used as
the input demography for \msprime simulations. We sampled the same
number of individuals from each population as used in the inference (Nama: 44,
Mende: 85, Gumuz: 23, combined Oromo/Amhara: 46, British: 91, and Vindija
Neanderthal: 1). We assumed a generation time of 29 years, a per-base mutation
rate of $1.25\times10^{-8}$, a per-base recombination rate of $1\times10^{-8}$
(corresponding to 1cM/Mb), and we simulated 100 replicates with sequence
lengths of 50Mb per replicate. We computed the conditional site frequency
spectrum (cSFS) conditioned on observing the derived variant in the Neanderthal
sample, and reconstructed gene genealogies using \Relate to obtain coalescence
rate trajectories and distributions of deep branches passing through the
Neanderthal branch. Each analysis followed the same procedure as described
above in the analysis of the real data.

@@

All supplementary inference results are provided in Demes format
\citep{Gower2022-yn} as a supplemental zipped archive.

@@

In the demographic inference results presented in the main text, we did not
impose a minimum threshold on inferred Nama ancestry (as estimated using
AMDIXTURE). Because our proposed models allow for post-divergence gene flow as
well as recent admixture events, we can jointly learn deeper history and recent
admixture dynamics. This is in contrast to methods that do not account for
ongoing migration or recent admixture events, for which un-admixed genomes (if
they even exist) are needed for unbiased estimates of earlier history.
Although the methods used in this paper (\texttt{moments-LD}) are able to infer
early and recent history jointly, we tested the robustness of the inferred
early history to varying ancestry proportions in the Nama population. To do
this, we used ADMIXTURE to cluster inferred ancestry components, noting that
ADMIXTURE-inferred ancestries are coarse and can be misleading about true
underlying admixture proportions. Nonetheless, we included Nama individuals
that exceeded either 90\% ro 99\% ancestry primarily shared among Nama
individuals.

For both thresholds, early history was robust in all tested models. Divergence
times and migration rates did not vary significantly from the original fits
that included all unrelated Nama individuals. The primary differences between
these fits and the original fits that did not impose an ancestry threshold were
the inferred admixture proportions of East African agriculturalists (2ka) and
Europeans (10 generations ago) in the Nama population. The East African
admixture proportion was reduced from $\approx 25\%$ to $20-22\%$, while the
European admixture proportion was reduced from $\approx 15\%$ to either $5-6\%$
(with a $90\%$ ancestry cutoff) or $3-5\%$ (with a $99\%$ ancestry cutoff). No
other parameters were strongly affected by subsetting the Nama individuals by
ADMIXTURE-inferred ancestry proportions. Because only the inferred proportions
of recent admixture in our model fits changed, we conclude that our inferences
of early history are robust to recent admixture, as long as that recent
admixture is accounted for in the demographic models.
